---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# An open-source framework for Rift Valley Fever forecasting

<!-- badges: start -->
[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License (for code): MIT](https://img.shields.io/badge/License (for code)-MIT-green.svg)](https://opensource.org/licenses/MIT)
[![License: CC0-1.0](https://img.shields.io/badge/License (for data)-CC0_1.0-lightgrey.svg)](http://creativecommons.org/publicdomain/zero/1.0/)
[![License: CC-BY-4.0](https://img.shields.io/badge/License (for text)-CC_BY_4.0-blue.svg)](http://creativecommons.org/publicdomain/zero/1.0/)
<!-- badges: end -->

### OpenRVFcast
EcoHealth Alliance's ongoing OpenRVFcast project is developing a generalizable, open-source modeling framework for predicting Rift Valley Fever (RVF) outbreaks in Africa, funded by the Wellcome Trust's climate-sensitive infectious disease [modeling initiative](https://wellcome.org/news/digital-tools-climate-sensitive-infectious-disease). We aim to integrate open data sets of climatic and vegetation data with internationally-reported outbreak data to build an modeling pipeline that can be adapted to varying local conditions in RVF-prone regions across the continent.

### Repository Structure and Reproducibility
-   `data/` contains downloaded and transformed data sources. These data are .gitignored and non-EHA users will need to download the data. 
-   `R/` contains functions used in this analysis. 
-   `reports/` contains literate code for  R Markdown reports generated in the analysis
-   `outputs/` contains compiled reports and figures.
-   This project uses the [{renv}](https://rstudio.github.io/renv/) framework to 
    record R package dependencies and versions. Packages and versions used are recorded in `renv.lock` and code used to manage dependencies is in `renv/` and other files in the root project directory.  On starting an R session in the working directory, run `renv::restore()` to install R package dependencies.
-   This project uses the [{targets}](https://wlandau.github.io/targets-manual/) 
    framework to organize build steps for analysis pipeline. The schematic figure below summarizes the steps. (The figure is generated using `mermaid.js` syntax and should display as a graph on GitHub.  It can also be viewed by pasting the code into <https://mermaid.live>.)
    
    
```{r, echo=FALSE, message = FALSE, results='asis'}
mer <- targets::tar_mermaid(targets_only = TRUE, outdated = FALSE, 
                            legend = FALSE, color = FALSE, 
                            exclude = c("readme", ends_with("_targets")))
cat(
  "```mermaid",
  mer[1], 
  #'Objects([""Objects""]) --- Functions>""Functions""]',
  'subgraph Project Workflow',
  mer[3:length(mer)],
  'linkStyle 0 stroke-width:0px;',
  "```",
  sep = "\n"
)
```

To run the pipeline, the user will need to adapt the `_targets.R` file to use their own object storage repository (we use AWS) and will need to supply keys in an `.env` file. 

For handling gridded binary weather data, this pipeline uses the [ecCodes](https://confluence.ecmwf.int/display/ECC) package which can be installed on OSX using `homebrew`:

```
brew install eccodes
```

and on Ubuntu using apt-get

```
sudo apt update
sudo apt install eccodes
```

EHA users: see the `stripts/` repository to be able to directly download data from AWS outside of the targets workflow. 

Follow the links for more information about:

- [`targets`](https://ecohealthalliance.github.io/eha-ma-handbook/3-projects.html#targets)
- [`renv`](https://ecohealthalliance.github.io/eha-ma-handbook/3-projects.html#package-management-with-renv)  
- [git-crypt](https://ecohealthalliance.github.io/eha-ma-handbook/16-encryption.html)
- [Reproducible workflows](https://github.com/ecohealthalliance/building-blocks-of-reproducibility)

