---
title: "initial-database-planning"
output: github_document
date: "2023-03-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(targets)
library(tidyverse)
library(arrow)
library(lubridate)
h <- here::here
```


#### Weather data (NASA)
Files are split by year and region
```{r}
ex_nasa_recorded_weather_local <- tar_read(nasa_recorded_weather_local, store = h("_targets"))[[1]]
ex_nasa_recorded_weather_local <- read_parquet(h(ex_nasa_recorded_weather_local))
weather <- ex_nasa_recorded_weather_local |>
  select(x = LON, y = LAT, date = YYYYMMDD, relative_humidity = RH2M, temperature = T2M, precipitation = PRECTOTCORR)
head(weather)
```

### Forecast data
Currently split by year
need to split by year, step, variable
```{r}
ex_ecmwf_forecasts_preprocessed_local <- tar_read(ecmwf_forecasts_preprocessed_local, store = h("_targets"))[[1]]
ex_ecmwf_forecasts_preprocessed_local <- read_parquet(h(ex_ecmwf_forecasts_preprocessed_local))
combos <- expand.grid(c("mean", "min", "max", "sd"), c("2m_dewpoint_temperature", "2m_temperature", "total_precipitation"), paste0("ahead_month_", 1:6)) 
vars <- paste(combos[, 1], combos[, 2], combos[, 3], sep = "_")
length(vars)
c("x", "y", vars, "days_til_first_forecast") 
# is it "how many days until 'month 1' begins?" or "how many days are we into `month 1`"?
# if it's the former, then maybe we are using month 2 from the previous forecast release as our "month 1" value?
```


#### NDVI (sentinel 2018-present)
Files are split by 10 day satellite period (all Africa in file)
Looks like the coverage is continuous and may overlap by a day
Note that this is s3a only. Need to confirm if we need s3b as well.
```{r}
ex_sentinel_ndvi <- tar_read(ndvi_local, store = h("_targets"))
ndvi_file_name <- ex_sentinel_ndvi[[1]]
ex_sentinel_ndvi <- terra::rast(h(ndvi_file_name))
ex_sentinel_ndvi <- as.data.frame(ex_sentinel_ndvi, xy = TRUE) |> as_tibble() |> slice(1:100)
start_date <- str_extract(ndvi_file_name, "(?<=_)(\\d{8})(?=T\\d{6})")
end_date <- str_extract(ndvi_file_name, "(?<=_)(\\d{8})(?=T\\d{6}_cache)")

ndvi <- crossing(tibble(date = seq.Date(from = ymd(start_date), to = ymd(end_date), by = "day")), ex_sentinel_ndvi)
head(ndvi)
```

#### Notes on CV and variable selection
https://arxiv.org/pdf/2303.07334.pdf
see `spatialsample` package
options: blocking, spatial clustering, leave one disc out

https://onlinelibrary.wiley.com/doi/10.1111/geb.13635

- Map accuracy estimates based on the relationship between the dissimilarity index (based on similarity of predictors in the holdout set to the training set) and the model performance

- The dissimilarity index is the normalized Euclidean distance to the nearest training data point in the multivariate predictor space, with predictors being scaled and weighted by their respective importance in the model (see Meyer & Pebesma, 2021, for more details on the calculation of the dissimilarity index). The area of applicability is then derived by applying a threshold to the dissimilarity index. The threshold is the (outlier-removed) maximum dissimilarity index of the training data derived via cross-validation.

- Reduce the number of predictors by spatial variable selection and compare the map accuracy and the area of applicability between the models using the full predictor set to models using the reduced predictor set in order to measure the benefits of spatial variable selection
